The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2016-03-06 oil_prices_obes.tex
corrected draft

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
Words: 5970
\end{abstract}

@@

Oil and gas supply, especially in difficult and expensive offshore areas, is the production of an often lengthy and complex chain of conditional investment decisions. Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled. In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability. For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval. Operating agreements are signed and subcontractors acquired for offshore construction and well development. Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase. All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process. \footnote{\citet{leffler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}

A large body of litterature exists on the many aspects of the complex chain of investment and production decisions made in the oil and gas industry. This article looks at the limited, but important topic of producing offshore fields, where a significant gap exists in the literature.

@@

Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.  the build-out and depletion phases of a field are conceptually unique and how oil companies operating offshore respond to oil prices can be expected to vary substantially.

The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price. The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments. Figure \ref{investment} shows that investments in producing fields makes up a large share of total investments, nearly dwarfing investments in searching, exploration and new build-outs. How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.

\begin{figure}
	\center
	\includegraphics[width=.9\textwidth]{figures/investment.png}
	\caption{The figure shows investments (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf. A large and growing share of investment goes into fields that are already producing as opposed to investments in search, exploration and build-out of new fields.}
	\label{investment}
\end{figure}

@@

But in practice, the question is not as clear cut.  Production in the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.  Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.  Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements. Higher oil prices may however still lead a producer to invest in increased capacity.  Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.

@@

However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.  The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector. \footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.

@@

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.  A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
	\caption{The top panel shows the production from the largest 10 oil fields, whose starting times are temporally correlated with each other.  The total production over time is bell-shaped, as shown by the bottom panel.}
	\label{oil_decline}
\end{figure}

The direction of this bias can be gleaned in figure \ref{oil_decline}.  High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000. These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production (\citep{osmundsen_is_2007}, \citep{aune_financial_2010}). However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.

@@

Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level which allowed oil supply to move in opposite direction of prices. However \citet{hamilton_oil_1983} showed that macroeconomic events could not explain movements in oil prices, and in turn rejecting the notion that prices contained all relevant information to oil supply. \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in emperically modeling oil production at the national level. The author used a hubbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, in turn using the residuals to model the effects of price on production in the lower 48 United States. The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved accurate up until the technological revolution of shale oil extraction.

@@

\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements. Instead, Adelman argued that cartel behavior of low cost producers likely stood behind the increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of the litterature.

The economics litterature on petroleum production tends to be overweighted towards studies of the US. \citet{watkins_world_1998} attempts to estimate supply functions for 41 countries by using methods from estimating the relationship between the price of undeveloped reserves and reserve additions over time. Based on the aggregated estimates, they found that fears of increasing global scarcity at the time were overblown. On the issue of endogeneity of price and supply, they make the somewhat dubious argument that for non-OPEC countries, price could be taken as exogenous since individual oil firms hedged (or "managed") price swings.

\citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added global reserves of increased oil prices and finds a small though statistically significant effect.  \citet{ramcharran_oil_2002} estimates a supply function for the total supply of oil from several OPEC countries based on data from 1973 to 1997.  The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.  However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt. \citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a countries oil industry. Though in the long run average, they find an elasticity of close to one.

On the surface, this article fits well into the previous econometrics litterature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces. However, most of the above articles focus on total long-term regional oil supply - which, as I mention above, is the result of a multiple chain of decisions from surveying of a potential play, exploration drilling, and production. This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields. While ambiguity exists if geophysical forces and scarcity is a significant factor that needs to be controlled for in the long term regional perspective, it is without question a dominant factor at the field level.

An expanding litterature looking at more microeconomic issues of oil supply exists. \citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration. The author finds that oil exploration firms in Texas approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.

@@

Most of the microeconometric articles analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity. Microeconomic studies of production from existing, operational fields are more sparse and make up a significant gap in the litterature where this article contributes.

Only a few papers study production decisions utilizing field-level data.  \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model. The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect. Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.  By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on onshore oil wells in California to estimate the effects of tax changes and price controls. The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.  These findings are over-all consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production. The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.  How producers react to a short-term tax change as opposed to a change in price also likely differs.

@@

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field. Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming. A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}

@@

The tax system is also designed to be neutral, that is a field that is profitable before tax should also be profitable after tax. This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct and carry forward losses, and deducting investment costs.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}

The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.  In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.  But this was mainly levied on the use and import of petroleum products.  In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.

@@

To simplify the analysis, I eliminate the smallest fields - those with less than 3 million SM3 of recoverable reserves.  These fields tend to have substantially different production profiles than the larger fields. However, these fields only account for around 0.5 \% of total recoverable reserves so this exclusion is unlikely to have any serious practical implications.

@@

\begin{table}
\begin{tabular}{lrrrrr}
\toprule
Percentiles &    5\% &    25\% &    50\% &     75\% &     95\% \\
                      &       &        &        &         &         \\
\midrule
In-place Oil, Mill SM3         &  5.90 &  34.00 &  75.80 &  157.00 &  636.70 \\
Est. Recoverable Oil           &  6.20 &  10.50 &  25.70 &   70.30 &  374.50 \\
Production Years, to 2015      &  1.79 &   9.17 &  16.04 &   22.43 &   35.63 \\
Tot. Cumulative Prod., to 2015, Mill SM3 &  3.61 &   8.94 &  22.94 &   57.46 &  359.04 \\
\bottomrule
\end{tabular}
\caption{Summary distribution statistics of field-level data. Columns represent percentiles of data. Excluds the Ekofisk field and fields with less than 5 million SM3 of recoverable oil.}
\label{field_summary}
\end{table}

@@

I do not, however, want to estimate smoothed curves individually for each field. While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.  Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.

A simple model can be written as in equation \ref{gam_price_eqn}.

\begin{equation}
\begin{split}
	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i}\\
	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
	  \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

@@

Another potential issue may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.  In particular, production profiles differ substantially by the size of the field. A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.  Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.

@@

The variables of interest in this article is the oil price and its lags, which I include as eight linear parametric terms in the model.  The idea of including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.  First, the field operator could be operating on the basis of some short-term extraction rule - choosing to pump out less at times of lower prices in order to pump out more at periods of high prices.

Alternatively, a change in price of oil can be seen as a lifting of a production constraint.  A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.  However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.  As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.

@@

\caption{Model results. Estimated standard coefficients are shown for the linear terms. Standard errors are in parenthesis. For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the terms significantly improve the fit of the model.}

@@

For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}. In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag. A negative coefficient for the eight lag is estimated in the 2-d model, however this is not significant in the model with random effects. Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/price_coefficents.png}
	\caption{Estimated coefficients from two GAM models: a model with random field effects and single smoothed function of production time, and another with a two dimensional smoothed function of production time and field size.  The points represent the point estimates, while the lines represent approximate 95\% confidence intervals. Significant positive coefficients are estimated on lags of between 1-4 years.  However, little to no effect is measured on the concurrent price term.}
	\label{price_coefficients}
\end{figure}

@@

The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}. The interaction effects, labeled $price lt:build$, are not estimated to be significantly different from zero. Given the small sample used to estimate these terms, this is not surprising. On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now subsantially larger in magnitude. In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase. This provides some evidence that the effect of price on production is indeed strongest in this latter stage of field development.

@@

The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 1 and 4 years.  Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.  Instead they are likely using storage or financial instruments to hedge short-term price movements. Changes in oil prices can rather be seen as a relaxing of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production.

@@

Another important caveat, is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may otherwise have happened later. Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives. First, there is the problem of censored data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.  More so, the estimates of both recoverable reserves and in-place oil are too uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.

@@

Policy implications arise for both energy industry regulators and financial policy makers. Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas. This is particularily important in the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices. \footnote{I thank an anonymous reviewer for pointing out this implication.}

>>>>

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
Words: 7000
\end{abstract}

@@

Oil and gas supply, especially in difficult and expensive offshore areas, is the product of an often lengthy and complex chain of conditional investment decisions. Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled. In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability. For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval. Operating agreements are signed and subcontractors acquired for offshore construction and well development. Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase. All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process. \footnote{\citet{lefler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}

A large body of economic litterature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry. This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.

@@

The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price. The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of total investment will be directed at capacity enhancement in producing fields rather than new field developments. Figure \ref{investment} shows that investments in producing fields makes up a large and growing share of total investment, nearly dwarfing investments in searching, exploration and new build-outs. How production from mature fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.

\begin{figure}
	\center
	\includegraphics[width=.9\textwidth]{figures/investment.png}
	\caption{The figure shows investment (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf. A large and growing share of total investment goes into fields that are already producing as opposed to investment in search, exploration and build-out of new fields.}
	\label{investment}
\end{figure}

@@

But in practice, the question is not as clear cut.  Production on the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.  Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.  Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements. Higher oil prices may still lead a producer to invest in increased capacity.  Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.

@@

However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.  The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector. \footnote{See \url{http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586}} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.

@@

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.  A modest effect can be detected at a lag of between 2 and 4 years of between 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
	\caption{The top panel shows the production from the largest 10 oil fields on the Norwegian Continental Shelf, whose starting times are temporally correlated with each other.  The total production over time is bell-shaped, as shown by the bottom panel.}
	\label{oil_decline}
\end{figure}

The direction of this bias can be gleaned in figure \ref{oil_decline}.  High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000. These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production \citep{osmundsen_is_2007}, \citep{aune_financial_2010}. However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.

@@

An extensive litterature exists that attempts to estimate the effect of oil prices on regional or global oil supply. Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level. \citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply. \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an emperical model of oil production in the US. The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production. The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.

@@

\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s. Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of litterature.

The economics litterature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted. \citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time. Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown.

\citet{farzin_impact_2001} attempts to estimate the price elasticity of added global reserves and finds a small positive elasticity.  \citet{ramcharran_oil_2002} estimates a total supply function of oil from several OPEC countries based on data from 1973 to 1997.  The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.  However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt. \citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a country's oil industry. Though in the long run, they find an average elasticity of close to one.

On the surface, this article fits well into the previous econometrics litterature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces. However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production. This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields. While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.

Microeconomic aspects of oil supply have more recently come to the fore. \citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration. The author finds that oil exploration firms in Texas approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.

@@

Most of the microeconometric articles analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity. Microeconomic studies of production from operational fields are more sparse. \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model. The major limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect. Methodologically, the paper also relies heavily on strict assumptions about the functional form of both the geo-engineering and economic aspects of their model.  By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on onshore oil wells in California to estimate the effects of tax changes and price controls. The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.  These findings are generally consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production. The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.  How producers react to a short-term tax change as opposed to a change in price also likely differs.

@@

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field. Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming. A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and is not present in the data set used for the analysis.}

@@

The tax system is also designed to be neutral, that is a field that is profitable before tax should also be profitable after tax. This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct investment costs and deduct and carry forward losses.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}

The over-all tax rate has been fairly constant through the history of Norwegian petroleum production, however several important changes related to the tax code have occurred.  In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.  But this was mainly levied on the use and import of petroleum products.  In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.

@@

To simplify the analysis, I eliminate the smallest fields - those with less than 5 million SM3 of recoverable reserves.  These fields tend to have substantially different production profiles than the larger fields. However, these fields only account for around 0.5 \% of total recoverable reserves.

@@

\begin{table}
\begin{tabular}{lrrrrr}
\toprule
Percentiles &    5\% &    25\% &    50\% &     75\% &     95\% \\
                      &       &        &        &         &         \\
\midrule
In-place Oil, Mill SM3         &  5.90 &  34.00 &  75.80 &  157.00 &  636.70 \\
Est. Recoverable Oil           &  6.20 &  10.50 &  25.70 &   70.30 &  374.50 \\
Production Years, to 2015      &  1.79 &   9.17 &  16.04 &   22.43 &   35.63 \\
Tot. Cumulative Prod., to 2015, Mill SM3 &  3.61 &   8.94 &  22.94 &   57.46 &  359.04 \\
\bottomrule
\end{tabular}
\caption{Summary distribution statistics of field-level data. Columns represent percentiles of data. Excludes the Ekofisk field and fields with less than 5 million SM3 of recoverable oil.}
\label{field_summary}
\end{table}

@@

I do not want to estimate smoothed curves individually for each field. While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.  Instead I want to estimate a general shape of the production profile for all fields. A simple model can be written as in equation \ref{gam_price_eqn}.

\begin{equation}
\begin{split}
	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i}\\
	 \quad & + \mathbf{\beta_2 oil\_price\_lags_t} + 
	 \beta_3 cost\_index_{t}  +  
	  \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

@@

Another potential issue may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.  In particular, production profiles differ substantially by the size of the field. A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.  Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i})$.

@@

The variables of interest in this article is the oil price and its lags, which I include as eight linear parametric terms in the model.  The reason for including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.  First, the field operator could be operating on the basis of some short-term extraction rule - choosing to pump out less at times of lower prices in order to pump out more at periods of high prices.

Alternatively, a change in the price of oil can be seen as a lifting of a production constraint.  A higher oil price means that added investments in production becomes attractive in order to either increase the total amount extracted from a field or to shift production forward.  However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.  As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.

@@

\caption{Model results. Estimated coefficients are shown for the linear terms. Standard errors are in parenthesis. For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the smooth terms significantly improve the fit of the model.}

@@

For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}. In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2st to 4th lags, with a peak at the 2nd lag. A negative coefficient for the eighth lag is estimated in the 2-d model, however this is not significant in the model with random effects. Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model. On the other hand, the block of terms for the 1st through 4th lags do significantly add to the fit of the model.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/price_coefficents.png}
	\caption{Estimated coefficients from two GAM models: a model with random field effects and single smoothed function of production time, and another with a two dimensional smoothed function of production time and field size.  The points represent the point estimates, while the lines represent approximate 95\% confidence intervals. Significant positive coefficients are estimated on lags of between 2-4 years.  However, little to no effect is measured on the concurrent price term.}
	\label{price_coefficients}
\end{figure}

@@

The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}. The interaction effects, labeled $price\_lt:build$, are not estimated to be significantly different from zero. Given the small sample used to estimate these terms, this is not surprising. On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now subsantially larger in magnitude. In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase. This provides some evidence that the effect of price on production is indeed strongest in this latter stage of field development.

@@

The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a modest effect is estimated  with a lag of between 2 and 4 years.  Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.  Instead they are likely using storage or financial instruments to hedge short-term price movements. Changes in oil prices can rather be seen as a relaxing of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production.

@@

Another important caveat is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may otherwise have happened later. Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives. First, there is the problem of censored data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.  More so, the estimates of both recoverable reserves and in-place oil are too uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.

@@

Policy implications arise for both energy industry regulators and financial policy makers. Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas. This is particularily important on the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices. \footnote{I thank an anonymous reviewer for pointing out this implication.}

### 2016-03-06 oil_prices_title.tex
corrected draft

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 1 to 4\% for a 10 dollar per barrel increase in the real price of oil.\\
\textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\
\textbf{Words}: 5970
\end{abstract}

>>>>

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.\\
\textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\
\textbf{Words}: 5970
\end{abstract}
